---
title: "Data wrangling with dplyr"
author: "Elisa Bayraktarov"
date: "19 June 2019"
output: html_document
---
## Code Objective
- discuss tidy data
- read data from online into R
- explore gapminder data with base-R functions
- wrangle gapminder data with dplyr tidyverse functions
- practice RStudio-GitHub workflow

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading the tidyverse
```{r packages}
library(tidyverse)

```

Load the data
```{r}
# loading the gapminder dataset from github
gapminder <-read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv")

head(gapminder)

summary(gapminder)

# shows the last 6 records
tail(gapminder)

# this shows you the structure of the data, the data is a data.frame
str(gapminder)

# Show me the name of the columns
names(gapminder)

# give me the number of columns:
ncol(gapminder)

# the variables e.g. countries are columns
# every raw is a unique observation

```

Look at individual dataframe variables
```{r}
# show the lifeExp column

gapminder$lifeExp
gapminder$continent

```

Learning the filter() function
```{r}

filter(gapminder, lifeExp < 30)

# filter for observations only where the country equals Mexico only; just look at Mexico and only for observations in year 2002
filter(gapminder, country == "Mexico", year == 2002)

x <- filter(gapminder, country == "Brazil", year >= 1987, year <= 2007)

mean(x$lifeExp)

```

Learning the select() function
```{r}

# select only these columns:
select(gapminder, year, country, lifeExp)

# drop some columns from the dataset;
select(gapminder, -continent, -gdpPercap, -pop)

```

Using filter() and slect together
```{r}

## instead of this...
gap_cambodia  <- filter(gapminder, country == "Cambodia")
gap_cambodia2 <- select(gap_cambodia, -continent, -lifeExp) 

## ...we can do this
gap_cambodia  <- gapminder %>% filter(country == "Cambodia")
gap_cambodia2 <- gap_cambodia %>% select(-continent, -lifeExp) 

```

The pipe operator %>% 
Ctrl + Shift + m
```{r}

gapminder %>% head()
head(gapminder)

# the pipe: take gapminder (on the left) and then give me the head (what's on the right)


## instead of this...
gap_cambodia  <- filter(gapminder, country == "Cambodia") %>% #and then
select(-continent, -lifeExp)


gap_cambodia  <- gapminder %>% 
  filter(country == "Cambodia") %>% #and then
  select(-continent, -lifeExp)

# We want year, pop and gdpPercap for Cambodia only. We use the pipe: Take the gapminder dataframe, and then filter for the country of Cambodia and then select all columns except for continent and lifeExp

gap_cambodia
```


Learn mutate() which adds a new column and calls it as you want it to call e.g. gdp
```{r}

gapminder %>%
  mutate(gdp = pop*gdpPercap)

# Take the dataframe gapminder and then make a new column called gdp which is the population times the gdpPercap

# Calculate the population in thousands for all Asian countries in the year 2007 and add it as a new column. Then show a new dataframe which has country, year and population in thousands

head(gapminder)

gapminder %>% 
  filter(continent =="Asia", year == 2007) %>% 
  mutate(population_in_thousands = pop/1000) %>% 
  select(country, year, population_in_thousands)

```

Learning the group_by function
```{r}
# The total pop for each continent in 2002 only
gapminder %>%
  filter(year == 2002) %>% 
  group_by(continent) %>% 
  mutate(cont_pop = sum(pop))
  


```

Add in summarise()
```{r}
# The total pop for each continent in 2002 only
gapminder %>%
  filter(year == 2002) %>% 
  group_by(continent) %>% 
  summarize(cont_pop = sum(pop))


```

```{r}
gapminder %>%
  group_by(continent,year) %>% 
  summarize(cont_pop = sum(pop)) %>% 
  ungroup() # Good practice to add ungroup after you do a grouping variable

gapminder %>%
  group_by(continent,year) %>% 
  summarize(cont_pop = sum(pop)) %>% 
  arrange(year) # arrange the data by ascending years

gapminder %>%
  group_by(continent,year) %>% 
  summarize(cont_pop = sum(pop)) %>% 
  arrange(-year) # arrange the data by descending years

```

What is the maximum GDP per continent across all years?
```{r}

gapminder %>%
  mutate(GDP = pop*gdpPercap) %>% 
  group_by(continent) %>%
  mutate(max_gdp = max(GDP)) %>% 
  filter(GDP ==max_gdp)

```


