---
title: "Graphics with ggplot2"
author: "Elisa Bayraktarov"
date: "18 June 2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

## Code objective
The objective of this code is to see how to configure and use GitHub. The second part of this code is to produce graphics with ggplot2

Don't forget, the shortcut for chunks is:
Ctrl + Alt + I

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loading packages, include=TRUE}
library(tidyverse)
library(ggplot2)

```

You can also read in data from webpages e.g. from GitHub, so get the data from GitHub:
```{r data}

ohi_data <-read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/OHI_global_data.csv")

head(ohi_data)
summary(ohi_data)

# the first part is the dataframe and the part after the $ is the column
table(ohi_data$georegion_one)
table(ohi_data$georegion_two)

# how big is the data (giving raws and columns)
dim(ohi_data)

```

Let's plot the data:
```{r start ggplot}

# First line of the ggplot is the dataframe and telling inside of the aes what the x and y axes are:
ggplot(data=ohi_data, aes(x=georegion_one, y=OHI_score))+
  geom_point()

ggplot(data=ohi_data, aes(x=georegion_one, y=OHI_score))+
  geom_jitter(width=0.1)

?geom_jitter
```

Let's do a boxplot:
```{r boxplot}
ggplot(data=ohi_data, aes(x=georegion_one))+
  geom_bar(width=0.9)

```

Let's do a Histogram:
```{r hist}

ggplot(data=ohi_data, aes(x=HDI))+
  geom_histogram()

```

Multiples geoms exploration:

```{r multiple geoms}

ohi_summary <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/OHI_scores_georegion_summary.csv")

# bar plot is produced with the same values as in the data
ggplot(data=ohi_summary, aes(x=georegions, y=OHI_score_average))+
  geom_bar(stat="identity")+
  geom_jitter(data = ohi_data, aes(x = georegion_one, y = OHI_score), width = 0.1)

# Are the next 2 arguments the same? - yes they are
ggplot(data = ohi_data, aes(y=OHI_score, x = HDI, color=georegion_one)) +
  geom_point()

# it's better to put the aes with teh geoms if you put many layers of plots on top of each other  
ggplot(data = ohi_data) +
  geom_point(aes(y = OHI_score, x = HDI, color=georegion_one))

# Here are some examples of aes arguments:  
# color color of lines/points
# fill color within polygons
# label name
# linetype type of line
# shape style of point
# alpha transparency (0-1)
# size size of shape

```


Adding a third varibale with aes
```{r}

ggplot(data=ohi_data, aes(x = OHI_score, y = HDI, size = coastal_pop))+
  geom_point()

ggplot(data=ohi_data, aes(x = OHI_score, y = HDI, color = coastal_pop))+
  geom_point()

ggplot(data=ohi_data, aes(x = OHI_score, y = HDI, alpha = coastal_pop))+
  geom_point()

```

Adding a third discreate variable with aes
```{r}

ggplot(data=ohi_data, aes(x = OHI_score, y = HDI, color = georegion_one))+
  geom_point()

ggplot(data=ohi_data, aes(x = OHI_score, y = HDI, shape = georegion_one))+
  geom_point()

```

Make prettier plots by using Themes
theme_bw() is just a pre-set theme of black and white
```{r}

ggplot(data=ohi_data, aes(x = OHI_score, y = HDI))+
  geom_point()+
  theme_bw()

source('https://raw.githubusercontent.com/OHI-Science/ohiprep/master/src/R/scatterTheme.txt')    

ggplot(data=ohi_data, aes(x = OHI_score, y = HDI))+
  geom_point()+
  scatterTheme


```

Change labels
```{r}

ggplot(data = ohi_data, aes(OHI_score, x = HDI, color = georegion_one))+
  geom_point()+
      labs(y = "OHI score, 2017",
       x = "Human Development Index",
       title = "Countries with high human development have more sustainable oceans",
      color = "Georegion") +  # if color doesn't work, use "fill"
     theme_bw()

```


```{r}
# install.packages("RColorBrewer")
library(RColorBrewer)
display.brewer.all()

my_palette <- brewer.pal(n=9, "YlOrRd")

ggplot(data = ohi_data, aes(x = OHI_score, y = OHI_trend, color = HDI)) + 
  geom_point(size =3) +
  scale_colour_gradientn(colors = my_palette)

ggplot(data = ohi_data, aes(x = OHI_score, y = OHI_trend, color = HDI)) + 
  geom_point(size =3) +
  scale_colour_gradientn(colors = c("blue", "green"))

# lets use a discrete color scale
my_palette <- brewer.pal(n=12, "Set3")

ggplot(data = ohi_data, aes(x = OHI_score, y = HDI, color = georegion_one)) + 
  geom_point(size = 3) +
  scale_color_manual(values = my_palette)

```


